MLH1 Results Outline

### 2020-04-08 doc/MatMeth.Rmd
checking all code chuncks after integrating brets comments

---
title: "Material Methods"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: yes
    toc_depth: '5'
  pdf_document:
    toc: yes
    toc_depth: '5'
  html_document:
    toc: yes
    toc_depth: '5'
    toc_float: yes
bibliography: refs_MatMeth.bib
---

@@

>>>>

---
title: "Material Methods"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: yes
    toc_depth: '5'
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '5'
  word_document:
    toc: yes
    toc_depth: '5'
bibliography: refs_MatMeth.bib
---

@@

```{r data.sets, echo=FALSE}

#Q1.real.short_mouse_table

#

ToDo; Bret's suggestion for moving thing to Mat and Methods

**For SC based metric**

- To account for confounding effects of sex chromosomes from pooled samples of bivalents, we considered a reduced dataset including only bivalents with SC lengths below the 2nd quartile (for SC length) in each cell. This dataset included the four or five shortest bivalents  and excluded the X bivalent in oocytes. A total of 678 -- `r sum(Q1.real.short_mouse_table$n_obs)` 'short' bivalents were isolated from 103 -- `r sum(  (Q1.real.short_mouse_table  %>% filter(sex=="female") )$ncells)` oocytes and 37 -- `r sum(  (Q1.real.short_mouse_table  %>% filter(sex=="male") )$ncells)` spermatocytes. Although this smaller dataset has decreased power, it offers a more comparable set of single bivalents to compare between the sexes.

- As an additional metric of chromatin compaction, we computed the total (summed) SC length of all bivalents for single cells, using the image analysis algorithm of [@wang_2019_sc]. Mouse means were calculated from cell-wide total SC lengths in 2,984 -- `r length(  (MLH1.merge.Skel %>% filter(species == "M.musculus") %>% filter(subsp != "Cast")  )$Random.Name)` out of 3,680 -- `r length(  (MLH1_data %>% filter(species == "M.musculus") %>% filter(subsp != "Cast")  )$Random.Name)` cells with MLH1 counts (Figure X).

**SC Length calq**

-  **Linear model M1 ** $$mouse \ average \ bivalent metric ~=~ subsp * strain + \varepsilon $$

- **Linear model M2 ** $$mouse \ average \ bivalent metric ~=~ strain + \varepsilon $$

- This metric is closer to measures of interference from linkage maps which are also removed from physical scales in that they measure frequencies of crossovers. Crossover interference is stronger in male specific linkage maps compared to female maps in a variety of species (ref).  **IFD^norm^**

### 2020-04-08 doc/Results_v2.rmd
checking all code chuncks after integrating brets comments

---
title: "MLH1 Results Outline"
date:
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '5'
    toc_float: yes
  pdf_document: default
  word_document:
    toc: yes
    toc_depth: '5'
bibliography: refs_Main_Report.bib
---

@@

skel -- takes the skeletonized midline of the bin objects
-->
# Main Figure

@@

Graphical comparisons between the two sexes reveal several patterns (Figure 1A). First, recombination rate differs between females and males in most strains. Second, the direction and magnitude of heterochiasmy varies among strains. Although the majority of strains show higher recombination rates in females (following the pattern in laboratory mice), two musculus strains and one molossinus strain exhibit male-biased heterochiasmy. Hence, relative recombination rates in the two sexes are evolutionarily labile.
Separately examining the mouse means of MLH1 foci per cell for each sex points to distinct patterns of variation (Figure 1B-C). Female recombination rates are evenly distributed around the sex-wide mean of approximately 25 MLH1 foci per cell (Figure 1B). In stark contrast, males separate more clearly into two groups of strains with high rates (near 30 MLH1 foci per cell) and low rates (near 23 MLH1 foci per cell) (Figure 1C).

@@

Separtely examining the mouse means of MLH1 foci per cell separately for each sex reveal distinct patterns of variation (Figure1 B and C). Female recombination rate means are evenly distributed around the sex-wide mean of approximately 25 MLH1 foci per cell (Figure1 B). In stark contrast, males specific means separate more clearly into two groups of strains with high genome-wide recombination rates (near 30 MLH1 foci per cell) and low genome-wide recombination rates, (near 23 MLH1 foci per cell) (Figure1 C).

@@

<add a transition sentence, about the expectations under different evolutionary scnarios>

@@

To account for confounding effects of sex chromosomes from pooled samples of bivalents, we considered a reduced dataset including only bivalents with SC lengths below the 2nd quartile (for SC length) in each cell. This dataset included the four or five shortest bivalents  and excluded the X bivalent in oocytes. A total of 678 -- `r sum(Q1.real.short_mouse_table$n_obs)` 'short' bivalents were isolated from 103 -- `r sum(  (Q1.real.short_mouse_table  %>% filter(sex=="female") )$ncells)` oocytes and 37 -- `r sum(  (Q1.real.short_mouse_table  %>% filter(sex=="male") )$ncells)` spermatocytes. Although this smaller dataset has decreased power, it offers a more comparable set of single bivalents to compare between the sexes.

@@

As an additional metric of chromatin compaction, we computed the total (summed) SC length of all bivalents for single cells, using the image analysis algorithm of [@wang_2019_sc].

Mouse means were calculated from cell-wide total SC lengths in 2,984 -- `r length(  (MLH1.merge.Skel %>% filter(species == "M.musculus") %>% filter(subsp != "Cast")  )$Random.Name)` out of 3,680 -- `r length(  (MLH1_data %>% filter(species == "M.musculus") %>% filter(subsp != "Cast")  )$Random.Name)` cells with MLH1 counts (Figure X). Females have significantly longer total SC than males in each strain tested separately (t-test; p < 0.05). That females have longer SCs is also supported by mixed models and general linear models with covariates, which identify sex as the most consistently significant effect (p < 0.05). Additionally, there are some significant subspecies and strain effects (p < 0.05), indicating that SC length has evolved among strains and subspecies.

@@

Q1 Summary, consequences of sex differences in the recombination landscape
Our results confirm sex differences in recombination landscapes described in several other species and we note that these differences have distinct consequences on the potential patterns of genetic variation resulting from the distinct recombination landscapes.
The terminal position of single foci and the greater distance between two foci on the same chromosome will have a consequence of larger sections of linked sites segregating together in male gametes compared to female gametes (Veller).

@@

There is no strong signal of sex differences in raw mean inter-focal distances (IFDraw) across the full set of strains. A marginally significant difference between the sexes (t-test; p = 0.07) is driven by one strain (t-test without domesticusG; p = 0.27). This result indicates that females and males exhibit a similar level of interference when it is measured in physical (SC) units. In contrast, males have significantly longer normalized mean inter-focal distances (IFDnorm) than females in seven out of eight strains (t-tests; p < 0.02), with only musculusKAZ showing no difference (t-test; p = 0.33). Examination of IFDnorm distributions indicates that female IFDnorm values are centered at approximately 50% and show a slight enrichment of low (<25%) values, whereas males are enriched for higher values.
Mixed models and general linear models of IFDnorm support the inference of stronger interference in males: sex is the most significant variable (LRT: p = 6.7410^{-14}). When interference is measured in physical SC units (IFDraw), the differences between sexes is low and only slightly significant (data not shown). 
In summary, controlling for differences in chromatin compaction (SC length) using IFDnorm indicates that interference is stronger in males, whereas consideration of IFDraw shows that the sexes exhibit a similar level of interference on the physical (SC) scale.
While the comparisons of the IFDnorm metrics can reveal more general recombination landscape patterns while controlling for the underlying differences in chromatin compaction and SC length. This metric is closer to measures of interference from linkage maps which are also removed from physical scales in that they measure frequencies of crossovers. Crossover interference is stronger in male specific linkage maps compared to female maps in a variety of species (ref).

--- not sure what below is from --

Crossover interference is one of the major determinates of the recombination landscape. It generates a distribution of evenly spaced crossover along chromosomes instead of a random and more uniform distribution. To quantifiy how crossover interference differs between the typical recombination landscapes of males and females, we isolated 1360 -- `r length(  (Curated_BivData.22 %>% filter(sex == "female") )$Obj.ID )` and 1272 -- `r length(  (Curated_BivData.22 %>% filter(sex == "male") )$Obj.ID )` 2CO bivalents in females and males respectively. Mean interfocal distances (IFD) were calculated from 42 -- `r length( unique( (Curated_BivData.22 %>% filter(sex == "female") )$mouse) )` female and 45 -- `r length( unique( (Curated_BivData.22 %>% filter(sex == "male") )$mouse) )` male mice. We examined both raw IFDraw and normalized by SC length (IFDnorm). The mean IFDraw measures of crossover interference as a mechanical force mediated through the SC. We compared the IFDraw to test the hypothesis of interference acting as a mechanical force transmitted through the SC. Previous measures found no sex difference in measures of IFDraw in mice [@de_boer_two_2006].

----

There is no strong signal of sex differences in raw mean inter-focal distances (IFDraw) across the full set of strains. A marginally significant difference between the sexes (t-test; p = 0.07 -- `r Q1.IFD.raw_FULL_t.test$p.value`) is driven by one strain (t-test without domesticus^G^; p = 0.27 -- `r Q1.IFD.raw_sansG_t.test$p.value`). The difference in mean IFDraw are slightly significant sexes from the full data set (t.test; **p = 0.07)**. However domesticusG might be an outlier (by having a larger than average degree of sexual dimorphism).

When the domesticusG observations are removed, the difference is not longer significant (t.test; p = 0.27) <already stated?>.

@@

Crossover interference is one of the major determinates of the recombination landscape. It generates a distribution of evenly spaced crossover along chromosomes instead of a random and more uniform distribution. To quantifiy how crossover interference differs between the typical recombination landscapes of males and females, we isolated `r length(  (Curated_BivData.22 %>% filter(sex == "female") )$Obj.ID )` and 
`r length(  (Curated_BivData.22 %>% filter(sex == "male") )$Obj.ID )` 2CO bivalents in females and males respectively. Mean interfocal distances (IFD) were calculated from `r length( unique( (Curated_BivData.22 %>% filter(sex == "female") )$mouse) )`  female and `r length( unique( (Curated_BivData.22 %>% filter(sex == "male") )$mouse) )` male mice. We examined both raw IFD^raw^ and normalized by SC length (IFD^norm^). The mean IFD^raw^ measures of crossover interference as a mechanical force mediated through the SC. We compared the IFD^raw^ to test the hypothesis of interference acting as a mechanical force transmitted through the SC. Previous measures found no sex difference in measures of IFD^raw^ in mice [@de_boer_two_2006].

The difference in mean IFD^raw^ are slightly significant sexes from the full data set (t.test; p = `r Q1.IFD.raw_FULL_t.test$p.value`). However _domesticus^G^_ might be an outlier (by having a larger than average degree of sexual dimorphism). When the _domesticus^G^_ observations are removed, the difference is not longer significant (t.test; p = `r Q1.IFD.raw_sansG_t.test$p.value`).  This indicates there is no general pattern of sex differences in crossover interference measured in physical (SC) units.

While the comparisons of the IFD^norm^ metrics can reveal more general recombination landscape patterns while controlling for the underlying differences in chromatin compaction and SC length. This metric is closer to measures of interference from linkage maps which are also removed from physical scales in that they measure frequencies of crossovers. Crossover interference is stronger in male specific linkage maps compared to female maps in a variety of species (ref).

The difference in mean IFD^norm^ are highly significant between sexes. The mean IFD^norm^ are significantly long in six of seven strains (MSM t.test; p-value = `r Q1.IFD.PER_t.test_MSM$p.value`), _musculus^SKIVE^_ p-value = `r Q1.IFD.PER_t.test_SKIVE$p.value`), _musculus^PWD^_ p-value = `r Q1.IFD.PER_t.test_PWD$p.value`), _molossinus^MSM^_ p-value =`r Q1.IFD.PER_t.test_MSM$p.value`), _domesticus^LEW^_ p-value = `r Q1.IFD.PER_t.test_LEW$p.value`), _domesticus^G^_ p-value = `r Q1.IFD.PER_t.test_G$p.value`), and _domesticus^WSB^_ (t.test; p-value =`r Q1.IFD.PER_t.test_WSB$p.value`) ). Only _musculus^KAZ^_ did not have a difference in mean IFD^norm^ (t.test; p-value = `r Q1.IFD.PER_t.test_KAZ$p.value`).

@@

In examining the IFD^norm^ distributions, we note that the female IFD^norm^ observations are centered at approximately 50% and have slight an enrichment of short (<25%) IFD^norm^ observations. Comparitively, the IDF^norm^ distributions in males are enriched for longer IFD^norm^.

@@

## Additional Determinants of Genome-wide Recombination Rate Evolution in Males

@@

<equations>

**Linear model M1**

$$mouse \ average \ bivalent \ metric ~=~ subsp * rand(strain) + \varepsilon $$

**Linear model M2**

mouse average bivalent metric = strain+??

$$mouse \ average \ bivalent \ metric ~=~ rand(strain) + \varepsilon $$

**Linear model M1 ** $$mouse \ average \ bivalent metric ~=~ subsp * strain + \varepsilon $$

**Linear model M2 ** $$mouse \ average \ bivalent metric ~=~ strain + \varepsilon $$

>>>>

---
title: "MLH1 Results Outline"
date: "null"
output:
  word_document:
    toc: yes
    toc_depth: '5'
  pdf_document: default
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '5'
    toc_float: yes
bibliography: refs_Main_Report.bib
---

@@

skel -- takes the skeletonized midline of the bin objects
-->

@@

Graphical comparisons between the two sexes reveal several patterns (Figure 1A). First, recombination rate differs between females and males in most strains. Second, the direction and magnitude of heterochiasmy varies among strains. Although the majority of strains show higher recombination rates in females (following the pattern in laboratory mice), two musculus strains and one molossinus strain exhibit male-biased heterochiasmy. Hence, relative recombination rates in the two sexes are evolutionarily labile.

@@

Separately examining the mouse means of MLH1 foci per cell for each sex points to distinct patterns of variation (Figure 1B-C). Female recombination rates are evenly distributed around the sex-wide mean of approximately 25 MLH1 foci per cell (Figure 1B). In stark contrast, males separate more clearly into two groups of strains with high rates (near 30 MLH1 foci per cell) and low rates (near 23 MLH1 foci per cell) (Figure 1C).

@@

@@

@@

Females have significantly longer total SC than males in each strain tested separately (t-test; p < 0.05). That females have longer SCs is also supported by mixed models and general linear models with covariates, which identify sex as the most consistently significant effect (p < 0.05). Additionally, there are some significant subspecies and strain effects (p < 0.05), indicating that SC length has evolved among strains and subspecies.

@@

@@

There is no strong signal of sex differences in raw mean inter-focal distances (IFDraw) across the full set of strains. A marginally significant difference between the sexes (t-test; p = 0.07) is driven by one strain (t-test without domesticusG; p = 0.27). This result indicates that females and males exhibit a similar level of interference when it is measured in physical (SC) units. In contrast, males have significantly longer normalized mean inter-focal distances (IFDnorm) than females in seven out of eight strains (t-tests; p < 0.02), with only musculusKAZ showing no difference (t-test; p = 0.33). Examination of IFDnorm distributions indicates that female IFDnorm values are centered at approximately 50% and show a slight enrichment of low (<25%) values, whereas males are enriched for higher values.
Mixed models and general linear models of IFDnorm support the inference of stronger interference in males: sex is the most significant variable (LRT: p = 6.7410^{-14}). When interference is measured in physical SC units (IFDraw), the differences between sexes is low and only slightly significant (data not shown). 
In summary, controlling for differences in chromatin compaction (SC length) using IFDnorm indicates that interference is stronger in males, whereas consideration of IFDraw shows that the sexes exhibit a similar level of interference on the physical (SC) scale.
While the comparisons of the IFDnorm metrics can reveal more general recombination landscape patterns while controlling for the underlying differences in chromatin compaction and SC length.

@@

@@

@@

# Additional Determinants of Genome-wide Recombination Rate Evolution in Males

@@
